Time Stretching of the GeV Emission of GRBs: Fermi LAT data vs geometrical model

### 2014-01-17 Paper/gammaRays.tex
Added better model overview
Drafted arguments for the main assumption of the model

Gamma Ray Bursts (GRBs) are among the most energetic events in the Universe, therefore they might provide new knowledge for particle physics. Recent observatories, specifically Fermi \cite{Ackermann:2012kna} and Swift \cite{Gehrels:2004gu} allowed for extensive study of these explosions \cite{Vianello:2013ela,Gehrels:2013xd}. These and previous studies led to several interesting results. For example, the total energy emitted in gamma-rays during a burst was found to be similar among different bursts within an order of magnitude \cite{Bloom:2003wy}. Temporal variations of spectra were also studied. The spectral lags were found between different low energy bands \cite{Yi:2005ht}. The very high energy radiation was discovered to be extended if compared to x-ray emission \cite{Vianello:2013ela,Lange:2013uh}. Moreover, the Large Area Telescope (LAT) of Fermi allows us to explore the spectral variations even between high energy bands like $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we'll call them low and high energy bands). In this paper we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

We take the GRB list from the Fermi-LAT catalog \cite{Ackermann:2013zfa}. Tables 2 and 4 in \cite{Ackermann:2013zfa} provide the time, durations and locations of the bursts, which we use to download the events data from the LAT Data Server\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}. We also download events during a long time interval before the burst, and use the technique introduced in \cite{Rubtsov:2011qq} to estimate background radiation in both energy bands.
Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts for study, for which at least $10$ photons were detected in the high energy band. This leaves $3$ of them: 080916C \cite{Tajima:2009az}, 090902B \cite{Abdo:2009pg} and 090926A \cite{Bregeon:2011bu}.

For these 3 bursts we compute both high and low energy CDFs. We subtract the low energy background estimate from the low energy CDF, and add the high energy background estimate to it. This makes the low energy CDF non-monotonous and, rigorously speaking, we cannot use the 2-sample Kolmogorov-Smirnov test on it. However, since there are much more high energy events than there are low energy ones, we can think of the low energy CDF as continuous. Therefore, this non-monotonicity will be negligible, and will not harm the KS-test results very much.

@@

We propose that this result can be explained by the curvature effects (that is the effects of the jet geometry). This effects were explored by multiple authors \cite{Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba}. However, these studies were only concerned with x-ray radiation, and, more importantly, they assumed that the distribution of radiation sources is homogeneous throughout the jet. On the contrary, we propose a different idea: that the highest energy radiators are concentrated near the axis of the burst, that is the burst opening angle depends on energy.

\begin{figure}[modelOverview]
        \centering
        \includegraphics[width=\textwidth]{modelOverview}
        \caption{Model Overview}
        \label{fig:modelOverview}
\end{figure}

@@

>>>>

Gamma Ray Bursts (GRBs) are among the most energetic events in the Universe, therefore they might provide new knowledge for particle physics. Recent observatories, specifically Fermi \cite{Ackermann:2012kna} and Swift \cite{Gehrels:2004gu} allowed for extensive study of these explosions \cite{Vianello:2013ela,Gehrels:2013xd}. These and previous studies led to several interesting results. For example, the total energy emitted in gamma-rays during a burst was found to be similar among different bursts within an order of magnitude \cite{Bloom:2003wy}. This suggests that most of the bursts must have similar energetics in their rest frame. Temporal variations of spectra were also studied. The spectral lags were found between different low energy bands \cite{Yi:2005ht}. The very high energy radiation was discovered to be extended if compared to x-ray emission \cite{Vianello:2013ela,Lange:2013uh}. Moreover, the Large Area Telescope (LAT) of Fermi allows us to explore the spectral variations even between high energy bands like $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we'll call them low and high energy bands). In this paper we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

We take the GRB list from the Fermi-LAT catalog \cite{Ackermann:2013zfa}. Tables 2 and 4 in \cite{Ackermann:2013zfa} provide the time, durations and locations of the bursts, which we use to download events data from the LAT Data Server\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}. We also download events during a long time interval before the burst, and use the technique introduced in \cite{Rubtsov:2011qq} to estimate background radiation in both energy bands.
Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts for study, for which at least $10$ photons were detected in the high energy band. This leaves $3$ of them: 080916C \cite{Tajima:2009az}, 090902B \cite{Abdo:2009pg} and 090926A \cite{Bregeon:2011bu}.

For these 3 bursts we compute both high and low energy distributions of photon times. We subtract the low energy background estimate from the low energy CDF, and add the high energy background estimate to it. This makes the low energy CDF non-monotonous and, rigorously speaking, we cannot use the 2-sample Kolmogorov-Smirnov test on it. However, since there is a much greater number of photons on lower energies, we can think of the low energy CDF as continuous. Therefore, this non-monotonicity will be negligible, and will not harm the KS-test results much.

@@

We propose that this result can be explained by the curvature effects (that is the effects of the jet geometry). These effects were explored by multiple authors \cite{Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba}. However, these studies were only concerned with x-ray radiation, and, more importantly, they assumed that the distribution of radiation sources is homogeneous throughout the jet. We propose a contrary idea: that the highest energy radiators are concentrated near the axis of the burst, that is the burst opening angle depends on energy. We cannot prove this assumption rigorously, but we have arguments supporting it.

First of all, there are around 750 GRBs detected by GBM, half of which were in the LAT field of view \cite{Vianello:2013ela}. However, only about 30 of them were detected by the LAT, and only 3 of them were bright in the high energy band. If we extrapolate the uniform jet model to very high energies, it means that these groups of bursts are internally different: some of them produce VHE radiation, and other do not. These differences are hard to explain given that burst energetics are similar \cite{Bloom:2003wy}. Nevertheless, these differences in burst counts can easily be explained by our model. In our model, the opening angle of low energy jet is much larger than the opening angle of a high energy jet. Therefore the most common scenario for an observable jet is that the off-axis angle of the observer is smaller than low energy jet opening angle, but larger than the opening angle of a high energy jet. Because of that, most of the bursts are only seen at low energies. The 3 bursts we study in this paper were seen from the lowest off-axis angles.

Secondly, consider the jet plasma right after emission from the central engine. There, two processes happen simultaneously:
\begin{enumerate}
	\item{Plasma particles near the jet boundary change directions, therefore increasing the jet opening angle.}
	\item{Plasma particles lose energy, therefore decreasing radiation frequency emitted from the jet.}
\end{enumerate}
We argue that these two processes are correlated, because they happen due to the same particle interaction processes. And since they are correlated, jets with large opening angles will have lower energies, which is our main assumption.

\begin{figure}[modelOverview]
        \centering
        \includegraphics[width=0.7\textwidth]{modelOverview}
        \caption{
 %       	Model Overview. Since the total path (first as a plasma particle, then as a photon) is longer for high energies particles, the high energy emission is stretched compared to low energy emission.
 			Here the red cone represents the propagation volume of low energy plasma, while the blue cone stands for high energy radiators. The green disk is the observer. $\theta_2$ and $\theta_1$ are opening angles of high and low energy jets, $\chi$ is an observer's off-axis angle. Every ``particle'' emitted from the central engine should go through two stages before being observer: traveling in plasma to the radiation point, and then traveling to the observer as photons, If $\chi < \theta_1$, then due to the relativistic beaming effect, the majority of observed photons is radiated near the point $R_1$, which lies on the line from the central engine to the observer. However, it might be different for the high energy radiation. If $\theta_2 < \chi$, then there is no high energy plasma near $R_1$, and ``particles'' will have to move near the jet axis first, and then turn at emission to the observer's direction. For each moment of emission, this path is longer, therefore the high energy emission will be stretched with respect to the low energies.
        }
        \label{fig:modelOverview}
\end{figure}

Finally, this assumption is able to explain the non-trivial stretching factor of the GRB 090926A. The detailed computations will be described in the following sections, however, you can see why the stretching is happening on the figure \ref{fig:modelOverview}.

@@

\subsection{Download and Process}
\subsection{Linear Component (background)}
\subsection{KS-test}
\subsection{Results}
